---
title: "Final Project: CYP-GUIDES Trial"
author: "Thomas Chia, Eugene Choo, and Grady Hollar"
date: 'Due: Fri July 12 | 1:40pm'
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
editor_options: 
  markdown:
    wrap: 72
---

**APMA 3150 \| Summer 2024 \| University of Virginia\
*"On our honor, we pledge that we have neither given nor received help
on this assignment." - Thomas Chia, \_\_, Grady Hollar.***

------------------------------------------------------------------------

### Global Package Imports

```{r}
library(tidyverse)
library(car)
library(data.table)
library(gridExtra)
library(corrplot)
```

### Data Initialization

```{r}
data <- read.csv("data/Dataset.csv")
medication_key <- read.csv("data/Medicationkey.csv")
```

```{r}
# Lets begin by taking a look into the datasets.
head(data)
print(colnames(data))

head(medication_key)
print(colnames(medication_key))
```

------------------------------------------------------------------------

## ANOVA

```{r}
data$GENDER <- as.factor(data$GENDER)
data$`RACE.ETHNICITY` <- as.factor(data$`RACE.ETHNICITY`)
data$Assignment <- as.factor(data$Assignment)
data$EMR <- as.factor(data$EMR)
data$`Therapeutic.Guidances` <- as.factor(data$`Therapeutic.Guidances`)

anova_model <- aov(LOS ~ Assignment + GENDER + `RACE.ETHNICITY`, data = data)
summary(anova_model)
```

**Understanding the results:** The Assignment variable has a P-value of
0.825 which is significantly higher than our $\alpha$ of 0.05. Therefore
it does not impact the length of stay. However, gender and ethnicity
both have statistically significant P-values, indicating that these
variables indeed do have an impact on the length of stay.

```{r}
# Boxplots
par(mfrow = c(1, 3))  # Set up the plotting area to have 1 row and 3 columns

# Boxplot for ASSIGNMENT
boxplot(LOS ~ Assignment, data = data, main = "LOS by ASSIGNMENT",
        xlab = "ASSIGNMENT", ylab = "Length of Stay (LOS)", col = "lightblue")

# Boxplot for GENDER
boxplot(LOS ~ GENDER, data = data, main = "LOS by GENDER",
        xlab = "GENDER", ylab = "Length of Stay (LOS)", col = "lightgreen")

# Boxplot for RACE.ETHNICITY
boxplot(LOS ~ RACE.ETHNICITY, data = data, main = "LOS by RACE.ETHNICITY",
        xlab = "RACE.ETHNICITY", ylab = "Length of Stay (LOS)", col = "lightcoral")
```

------------------------------------------------------------------------

## Regression Analysis

### Setup

Reset the data

```{r}
data <- read.csv("data/Dataset.csv")
medication_key <- read.csv("data/Medicationkey.csv")
```

We start by converting the relevant columns into factors and changing
the medicine counts to indicator variables.

```{r}
# convert categorical variables to factors
data$GENDER <- as.factor(data$GENDER)
levels(data$GENDER) <- c("Female", "Male")
data$RACE.ETHNICITY <- as.factor(data$RACE.ETHNICITY)
levels(data$RACE.ETHNICITY) <- c("Black", "Latino", "Other/Unspecified", "White")
diagnosis_factor <- as.factor(data$Diagnosis)
data$Diagnosis <- diagnosis_factor
data$MD <- as.factor(data$MD)
data$Assignment <- as.factor(data$Assignment)
data$EMR <- as.factor(data$EMR)
data$Therapeutic.Guidances <- as.factor(data$Therapeutic.Guidances)


# change medicines to indicator variables rather than total counts (yes/no did recieve this medicine during hospital stay)
for (i in seq(0:29)) {
  data[i+10] <- as.numeric(data[i+10] > 0)
}
```

We aim to analyze the data by performing linear regressions to predict
the length of stay of a patient. We take two different approaches.
First, by considering the entire data set, and secondly by isolating the
most common diagnosis(es) and analyzing them separately. The reason we
do this is to try and account for the differences in length of stay
across different diagnoses, as can be seen in the plot below.

```{r}
ggplot(data, aes(x=Diagnosis, y=LOS))+
  geom_boxplot()+
  scale_x_discrete(labels=seq(1:32))
```

Before performing the regression, we examine the distribution of the
response variable.

```{r}
ggplot(data, aes(x=LOS))+
  geom_histogram(color="black", fill="#5e94ad")+
  labs(x="Length of Stay (days)", y="Frequency")
```

We see that the data is heavily skewed. To account for this, we perform
a log transformation on the data.

```{r}
data$LOS <- log(data$LOS)
```

Next, we need to actually find the most common diagnoses.

```{r}
ggplot(data, aes(x=Diagnosis))+
  geom_bar(color="black", fill="#5e94ad")+
  scale_x_discrete(labels=seq(1:32))
```

Using the above we collapse the data into the four most common
diagnoses, 21, 7,21 and 28.

```{r}
# get the top 4 most common diagnosis
diagnosis_A <- levels(diagnosis_factor)[21]
diagnosis_B <- levels(diagnosis_factor)[7]
diagnosis_C <- levels(diagnosis_factor)[22]
diagnosis_D <- levels(diagnosis_factor)[28]

diagnosis_A; diagnosis_B; diagnosis_C; diagnosis_D

data_A <- data[data$Diagnosis==diagnosis_A,]
data_B <- data[data$Diagnosis==diagnosis_B,]
data_C <- data[data$Diagnosis==diagnosis_C,]
data_D <- data[data$Diagnosis==diagnosis_D,]

# combine all 4 into a single data frame, for use later
diagnoses_combined <- data[data$Diagnosis==diagnosis_A | data$Diagnosis==diagnosis_B | data$Diagnosis==diagnosis_C | data$Diagnosis==diagnosis_D,]
```

### LOS Modeled on Diagnosis and Treatment Plan

We start by trying a model based on a patient's diagnosis and treatment
plan. We use backwards elimination examining $R^2$ in an attempt to
prune the model. First, we start with the entire data set.

```{r}
# level 1
los_vs_treatment_init <- lm(LOS ~ Diagnosis + Assignment + Therapeutic.Guidances, data=data)
summary(los_vs_treatment_init)$adj.r.squared

# level 2
print("level 2")
los_vs_treatment_l2A <- lm(LOS ~ Diagnosis + Assignment, data=data)
summary(los_vs_treatment_l2A)$adj.r.squared # winning model
los_vs_treatment_l2B <- lm(LOS ~ Diagnosis +Therapeutic.Guidances, data=data)
summary(los_vs_treatment_l2B)$adj.r.squared
los_vs_treatment_l2C <- lm(LOS ~ Assignment + Therapeutic.Guidances, data=data)
summary(los_vs_treatment_l2C)$adj.r.squared

# level 3
print("level 3")
los_vs_treatment_l3A <- lm(LOS ~ Assignment, data=data)
summary(los_vs_treatment_l3A)$adj.r.squared
los_vs_treatment_l3B <- lm(LOS ~ Diagnosis, data=data)
summary(los_vs_treatment_l3B)$adj.r.squared

# winning model
print("winning model:")
summary(los_vs_treatment_l3B)
```

Now we do the same but only within the most common diagnosis, MDD,
Recurrent, Severe Without Psychotic Features. Obviously, we remove
`Diagnosis` from the model.

```{r}
# level 1
los_vs_treatmentA_init <- lm(LOS ~ Assignment + Therapeutic.Guidances, data=data_A)
summary(los_vs_treatment_init)$adj.r.squared

# level 2
print("level 2")
los_vs_treatmentA_l2A <- lm(LOS ~ Assignment, data=data_A)
summary(los_vs_treatmentA_l2A)$adj.r.squared
los_vs_treatmentA_l2B <- lm(LOS ~ Therapeutic.Guidances, data=data_A)
summary(los_vs_treatmentA_l2B)$adj.r.squared 

# no improvement at level 2

print("winning model:")
summary(los_vs_treatmentA_init)
```

### LOS Modeled on Demographics

In the ANOVA test, we saw that race and gender both had significant
affects on the length of stay. Let's investigate this further by
performing the same process as above in an attempt to fit a regression
model based on patient demographics. Again, we start with the entire
data set.

```{r}
# level 1
los_fit_l1 <- lm(LOS ~ GENDER + AGE + RACE.ETHNICITY, data=data)
summary(los_fit_l1)$adj.r.squared

# level 2
print("level 2")
los_fit_l2_A <- lm(LOS ~ GENDER + AGE , data=data)
summary(los_fit_l2_A)$adj.r.squared
los_fit_l2_B <- lm(LOS ~ GENDER + RACE.ETHNICITY, data=data)
summary(los_fit_l2_B)$adj.r.squared
los_fit_l2_C <- lm(LOS ~ AGE + RACE.ETHNICITY, data=data)
summary(los_fit_l2_C)$adj.r.squared

# no improvement at level 2

# winning model
print("winning model:")
summary(los_fit_l1)
```

Now within just the most common diagnosis.

```{r}
# level 1
los_vs_demographicsA_init <- lm(LOS ~ GENDER + AGE + RACE.ETHNICITY, data=data_A)
summary(los_vs_demographicsA_init)$adj.r.squared

# level 2
print("level 2")
los_vs_demographicsA_l2A <- lm(LOS ~ GENDER + AGE, data=data_A)
summary(los_vs_demographicsA_l2A)$adj.r.squared
los_vs_demographicsA_l2B <- lm(LOS ~ GENDER + RACE.ETHNICITY, data=data_A)
summary(los_vs_demographicsA_l2B)$adj.r.squared
los_vs_demographicsA_l2C <- lm(LOS ~ AGE + RACE.ETHNICITY, data=data_A)
summary(los_vs_demographicsA_l2C)$adj.r.squared

# no improvement at level 2

# winning model
summary(los_vs_demographicsA_init)
```

### Understanding the Results

Unfortunately, all of our regression models resulted in extremely low
$R^2$ values, meaning none of the models hold any meaningful predictive
power—even with a logarithmic transform and an attempt to hone in on a
single diagnosis. We conclude that there is simply no linear
relationship between the length of stay and the predictors.

However, although the models themselves are not predictive in nature, we
do see that some of the generated coefficients are significant. In the
diagnosis/treatment model on the entire data set, both "MDD, recurrent
episodes" and "Unspecified Depressive Disorder" had positive
coefficients that were significant at the 0.05 level. Furthermore, in
the demographic models, the "Male", "AGE", and "White" all have positive
significant coefficients. Though the relationships might not be linear,
the significance of the coefficients suggest there may be a positive
relationship between length of stay and these two diagnoses and three
demographics.

### More Failed Tests

Can mention in sideshow if space/time provides. Can easily skip over if
wanted.

#### Simple Linear Regression Between LOS and Age

First the entire dataset.

```{r}
los_age_lm <- lm(LOS ~ AGE, data=data)
summary(los_age_lm)

ggplot(data, aes(x=AGE, y=log(LOS),))+
  geom_point()+
  geom_smooth(method=lm, se=FALSE)
```

Now we look at the top 4 most common diagnoses separately.

```{r}
los_age_lmA <- lm(LOS ~ AGE, data=data_A)
summary(los_age_lmA)$coefficients
summary(los_age_lmA)$r.squared

los_age_lmB <- lm(LOS ~ AGE, data=data_B)
summary(los_age_lmB)$coefficients
summary(los_age_lmB)$r.squared

los_age_lmC <- lm(LOS ~ AGE, data=data_C)
summary(los_age_lmC)$coefficients
summary(los_age_lmC)$r.squared

los_age_lmD <- lm(LOS ~ AGE, data=data_D)
summary(los_age_lmD)$coefficients
summary(los_age_lmD)$r.squared

ggplot(diagnoses_combined, aes(x=AGE, y=log(LOS), color=Diagnosis))+
  geom_point()+
  geom_smooth(method=lm, se=FALSE)
```

Again, terrible $R^2$ values, but significant coefficients. Once again
suggesting a non-linear relationship between the variables.

#### Logistic Regression on RAR

We investigate the readmission rate, performing a logistic regression
based off of the patients' diagnoses and treatment plans. Start by
splitting the data.

```{r}
# 80:20 train:test split ratio
train_split <- data[1:1200,]
test_split <- data[1201:1500,]
```

Now we build the model and perform predictions.

```{r}
rar_treatment_glm <- glm(RAR ~ Diagnosis + Assignment + Therapeutic.Guidances, data=data)
rar_pred <- as.numeric(predict(rar_treatment_glm, test_split, type="response") > 0.3)
conf_matrix <- table(Predicted = rar_pred, Actual = test_split$RAR)
conf_matrix
```

Lets also try a model based off of patient demographics

```{r}
rar_demographic_glm <- glm(RAR ~ GENDER + AGE + RACE.ETHNICITY, data=data)
rar_pred <- as.numeric(predict(rar_demographic_glm, test_split, type="response") > 0.3)
conf_matrix <- table(Predicted = rar_pred, Actual = test_split$RAR)
conf_matrix
```

Both models always predict no readmission for every single patient, even
with a threshold value of 0.3. The models, once again, have little to no
predictive power.

------------------------------------------------------------------------

## Paired T-Test

------------------------------------------------------------------------

## Propensity Score Matching
